Compression-based inference on graph data

### 2013-04-30 Bastards/src/bastards.tex
...

We also show that if our source has strictly more computational power than our resource bounded Kolmogorov complexity, there is always a non-zero probability of seeing a compressible string and colncluding that it is incompressible.

@@

All logartihms are base $2$.

@@

Another definition of the universal distribution is as $2^{-K^t(x)}$. In the non-bounded case the two definitions are equal up to a mulitplicative constant. In the recource bounded variant, we can only be certain of the following inequality:
\begin{align*}
2^{-K^t(x)} &\leq \sum_{p\;:\;u^t(p) = x} 2^{-|p|}  \\
2^{-K^t(x)} &\leq m^t(x) \\
-K^t(x) &\leq \log m^t(x) \\ 
K^t(x) &\geq \log m^t(x)
\end{align*}

@@

\begin{proof}
By definition ref{def:domination}, we have 
\begin{align}
\forall {c \exists x : K^t(x) \geq - \log p_q(x) + c} \label{line:non-invariance}
\end{align}

>>>>

We also show that if our source has strictly more computational power than our resource bounded Kolmogorov complexity, there is always a non-zero probability of seeing a compressible string and concluding that it is incompressible.

@@

All logarithms are base $2$.

@@

Another definition of the universal distribution is as $2^{-K^t(x)}$. In the non-bounded case the two definitions are equal up to a mulitplicative constant. In the recource bounded variant, we can only be certain of the following inequality:
\begin{align*}
2^{-K^t(x)} &\leq \sum_{p\;:\;u^t(p) = x} 2^{-|p|}  \\
2^{-K^t(x)} &\leq m^t(x) \\
-K^t(x) &\leq \log m^t(x) \\ 
K^t(x) &\geq -\log m^t(x)
\end{align*}

@@

\begin{proof}
By definition \ref{dfn:domination}, we have 
\begin{align}
\forall {c \exists x : K^t(x) \geq - \log p_q(x) + c} \label{line:non-invariance}
\end{align}

### 2013-04-30 Benelearn/src/benelearn.tex
...

\usepackage[accepted]{style/benelearn2013}
%\usepackage{charter}
\usepackage{eulervm}
%\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{subcaption}
\usepackage{float}
\usepackage{style/mlapa}

@@

\begin{abstract}
We investigate the use of compression-based learning on graph data. General purpose compressors operate on bitstrings or other sequential representations. Since a single graph can be represented sequentially in many ways, we study the effect of this projection. Using Normalized Compression Distance (NCD), we test a sequential compressor versus a native graph compressor. We use both synthetic, randomly generated graphs and real-life datasets.  We conclude that, even under adverse circumstances, sequential representations contain enough structure for shallow algorithms to perform inference successfully. Algorithms that operate directly on the graph representation require a considerable increase in resources, but do allow for an increase in performance also.
\end{abstract}

The mantra of compression-based inference is that \emph{compression equals learning}. If we can compress data, we must have found some structure, we must have learned something. Conversely, if we have learned something about data, we should be able to use that knowledge to represent our data more succinctly. \footnote{Other features commonly associated with learning, such as generalization will, so the adherents of compression-based learning argue, be optimized along with the optimization of compression. See, for instance \cite{grunwald2005tutorial,grunwald2007minimum}.}

@@

We can now say, that with respect to $U$, there must be a minimal description for any given data:
\[
K_U(x) = \min\left \{|y| : U(y) = x\right\} 
\]

The result that makes Kolmogorov Complexity a useful measure of information content is that $K_U(\cdot)$ is only marginally dependent on the choice of $U$. If we suppose that there is another universal description method, $V$, we might ask what the expected difference is between $K_U(x)$ and $K_U(v)$. Let $k_V(x)$ be the shortest program for $x$ on $V$. Since, $U$ is universal, we know that it can compute $k_v$ by simulating $V$. This simulation is a program for $x$ on $U$ which is a bound for the shortest program on $U$:
\begin{align*}
K_U(x) 	&\leq |\langle T_v, k_v(x)\rangle|\\ 
		&=  |k_v(x)| + |v| + p(v) \\
		&=  K_V(x) + O(1)
\end{align*}

@@

The second issue is one of scale. If two strings of a million bits differ by $1000$ bits, we might consider them quite similar, whereas two strings of $1000$ bits that differ by that amount could not be more different.\footnote{Note that this is only an intuitive example. If two strings differ in exactly every bit,a very short program transforms one into the other, so by NID, they are very similar.} Therefore, we would like to take the length of the strings into account. This gives us the Normalized information Distance (NID)

@@

When our data is represented as a graph, rather than a string, we replace the notion of concatenation strings by concatenation of graphs. That is, we combine the graphs $x$ and $y$ into a single (disconnected) graph.

@@

An important and subtle concern is the ordering of nodes in the sequential representation of our graphs. This issue is detailed very well in \cite{kang2011beyond}. As shown, there are various algorithms to determine node orderings that bring out a lot of the graph's inherent structure in the adjacency matrix, allowing a sequential compressor to use it. We can use substantial resources to find a good ordering of nodes to improve the performance of the sequential compressor, but if we did, we could no longer consider it a `shallow model'.

Since we are testing the capacity of the general purpose compressor to perform inference despite the problem of isomorphism, we will present it with a worst case scenario. We will use a random ordering of nodes for all graphs. If the general purpose compressor still outperforms the random baseline under these conditions, it will tell us that it is, at least in part, resistant to the problem of isomorphism.

@@

The first is the classic Erd\H{o}s-R\'enyi (ER) model, where a uniform random choice is made from all graphs with $n$ nodes and $m$ links. The second is the Barab\'asi-Albert (BA) model \cite{albert2002statistical}, which grows a graph from a set of $n_0$ unconnected nodes, one node at a time, connecting each new node to $k$ distinct existing nodes where the probability that a given existing node is chosen for a connection is its degree, divided by the sum of the degrees of all nodes. Thus, under the BA model the more links a node has, the higher the probability that it will accrue even more. This effect causes the degree distribution of a BA network to form a scale free distribution.

@@

To assess the performance of the clustering we label the clusters so that the accuracy is maximized (essentially assigning the optimal labels). Clearly, this would be cheating when testing a classifier, but since we are onnly interested in the clustering aspect, it gives us a straightforward performance measure.

@@

Subdue \cite{ketkar2005subdue} is an algorithm for finding frequent subgraphs in graph data. The algorithm searches through all possible subgraphs, searching for the  one that maximally compresses the data. The algorithm is essentially a beam search through the space of subgraphs. If consists of three main routines:

\begin{description}
\item[Subgraph matching] This is essentially an algorithm for finding the occurrences of a given subgraph in a graph. The algorithm used is detailed in \cite{bunke1983inexact}. Since this is essentially a semi-exhaustive search for the solution to the NP-Hard problem of subgraph isomorphism, the matching can only be solved for very small subgraphs. Unfortunately, weven with subgraphs of four or five nodes, the matching is too slow in combination with the number of times it is executed to calculate a full distance matrix. To combat this issue, we remove all but the first $b_{\mbox{inner}}$ elements from the search queue after it is sorted at ech iteration, effectively turning the algorithm into a beam search.
\item[MDL Scoring]
This routine takes a subgraph, finds its occurrences in the data by the previous routine and replaces these occurrences with a single node with a new label, which we call a \emph{symbol node}. All links that connected to nodes in the subgraph are now connected to the symbol node. The links connecting to the symbol node are annotated with the indices of the nodes in the subgraph they were connected to. Storing the subgraph once together with the reduced data graph is enough to reconstruct the original data, and, if the subgraph occurs often, will lead to a significant compression.
\item[Subgraph search]
This is the `outer loop' of the algorithm. Starting with a graph of a single link beteen two nodes, it searches through the space of all connected graphs by extending each current candidate by one link at a time (possibly by adding a new node as well). The buffer of current candidates is sorted by MDL score, and the candidate with the highest score is extended to create new candidates. At each iteration all but the top $b_{\mbox{outer}}$ candidates are removed, turning the search into a beam search.
\end{description}

@@

	\hspace*{5mm}\textit{sort $S$ by score($s', G$)} for $s' \in S$\\
	\hspace*{5mm}\textit{remove all but the first $b$ elements of $S$} \\
	\\
	\textbf{function score} $(s', G)$\\
	\hspace*{5mm} \textit{replace occurrences of $s'$ in $G$ with node $N$}\\
	\hspace*{5mm} \textit{annotate links to $N$ with the nodes in $s'$}\\
	\hspace*{5mm} \textbf{return} \textit{\# of bits to store the edited $G$ and $s'$ }\\
}
	\caption{Pseudocode for the Subdue algorithm}

@@

Table \ref{table:synthetic-large} shows the results on randomly generated graphs. Table \ref{table:real-life} shows the results for subgraphs sample from real-life datasets.

@@

As for the graph-compressors, we see a small improvement relative to the sequential compressors for a strong increase in computational resources. Subdue as used in this paper is a relatively simple compressor, which isolates only a single subgraph for compression and we tested it only at modest parameters. The publications surrounding Subdue offer much more complex solutions (such as the induction of graph grammars \cite{jonyer2004mdl}). To investigate the promise of these models as compressors further, it will be necessary to investigate both parallelized versions of these algorithms and a relaxation of the exhaustive nature of their components.

@@

The method of coding data is always a sensitive point in compression-based learning. The precise choices made in translating the data to a string of bits can greatly affect which patterns are picked up, or ignored by the subsequent inference procedure. Here we detail the exact procedure used to encode the graphs in both compressors.

\subsection*{GZIP}
We take half of the adjacency matrix ($(n^2 + n)/2$ bits), and store it in flattened form as an array of java byte primitives, together with the size of the string (since the stored representation is padded to a multiple of eight). We then serialize this representation into a java GZIPOutputStream. To make our implementation generic, we assume that the nodes and links are labeled, and serialize the labels in a fixed order after the adjacency matrix. In the graphs mentioned in this paper, there is a single label assigned to all elements, so inference shouldn;t be influenced.

@@

To store a plain graph, we follow roughly the coding strategy outlined in \cite{holder1994substructure}.

We first store the number of nodes, and the maximum number of neighbours for a node in the graph $n_{\mbox{max}}$. We then store the adjacency matrix row by row.

@@

We then account for the possibility of multiple links between node pairs. Since we expect this to be unlikely, we want to ensure that we do not waste bits coding for this occasion if it does not occur. We store the pairs for which there are multiple links as a list consisting of two numbers per element: the first is the index of the pair in a canonical ordering, the second is the number of extra links that exist for that pair (in prefix-coding). We use a single bit to indicate whether this list will follow, and we reserver a codeword to signal the end of the list.

After this, all that remains is to encode the labels of the nodes and links. We assume that the sender and receiver in our coding scheme posess a codebook that is efficient for the data given, so that if a label $l$ occurs with frequency $\#l$, we can encode it in $-\log \frac{\#l}{\sum_k \#k}$ bits.

We assume that there is some canonical ordering among the nodes and links, and store them as a stream. Since the number of labels is known from the adjacency matrix, this code is self-delimiting.

@@

Finally, we must store, for every link leading into a symbol node, which node in the substructure it should connect to to restore the original data. Since we know the graph and all the symbol nodes, we can presume there to be a canonical ordering in these links, and we use $\log n_s$ bits for each, where $n_s$ is the size of the substructure in nodes.

>>>>

% \usepackage{natbib}

\usepackage[accepted]{style/benelearn2013}
%\usepackage{charter}
\usepackage{eulervm}
%\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{subcaption}
\usepackage{float}

\usepackage{style/mlapa}

@@

\begin{abstract}
We investigate the use of compression-based learning on graph data. General purpose compressors operate on bitstrings or other sequential representations. A single graph can be represented sequentially in many ways, which may influence the performance of sequential compressors. Using Normalized Compression Distance (NCD), we test a sequential compressor versus a native graph compressor. We use both synthetic, randomly generated graphs and real-life datasets.  We conclude that, even under adverse circumstances, sequential representations contain enough structure for shallow algorithms to perform inference successfully. Algorithms that operate directly on the graph representation likely require a considerable increase in resources, but do allow for an increase in performance also.
\end{abstract}

The mantra of compression-based inference is that \emph{compression equals learning}. If we can compress data, we must have found some structure, we must have learned something. Conversely, if we have learned something about data, we should be able to use that knowledge to represent our data more succinctly. \footnote{Other features commonly associated with learning, such as generalization will, so the adherents of compression-based learning argue, be optimized along with the optimization of compression. See, for instance \npcite{grunwald2005tutorial,grunwald2007minimum}.}

@@

We can now say that with respect to $U$, there must be a minimal description for any given data:
\[
K_U(x) = \min\left \{|y| : U(y) = x\right\} 
\]

The result that makes Kolmogorov complexity a useful measure of information content is that $K_U(\cdot)$ is only marginally dependent on the choice of $U$. If we suppose that there is another universal description method, $V$, we might ask what the expected difference is between $K_U(x)$ and $K_U(v)$. Let $k_V(x)$ be the shortest program for $x$ on $V$. Since, $U$ is universal, we know that it can compute $k_v$ by simulating $V$. This simulation is a program for $x$ on $U$ which is a bound for the shortest program on $U$:
\begin{align*}
K_U(x) 	&\leq |\langle T_v, k_v(x)\rangle|\\ 
		&=  |k_v(x)| + |v| + p(v) \\
		&=  K_V(x) + O(1)
\end{align*}

@@

The second issue is one of scale. If two strings of a million bits differ by $1000$ bits, we might consider them quite similar, whereas two strings of $1000$ bits that differ by that amount could not be more different.\footnote{Note that this is only an intuitive example. If two strings differ in exactly every bit, a very short program transforms one into the other, so by NID, they are very similar.} Therefore, we would like to take the length of the strings into account. This gives us the Normalized information Distance (NID)

@@

When our data is represented as a graph, rather than a string, we replace the notion of concatenation strings by of concatenation of graphs. That is, we combine the graphs $x$ and $y$ into a single (disconnected) graph.

@@

An important and subtle concern is the ordering of nodes in the sequential representation of our graphs. This issue is detailed very well in \cite{kang2011beyond}. As shown, there are various algorithms to determine node orderings that bring out a lot of the graph's inherent structure in the adjacency matrix, allowing a sequential compressor to use it. We could use substantial resources to find a good ordering of nodes to improve the performance of the sequential compressor. I we did, however, we might lose the ability to consider the sequential compressor a shallow model. To maintain the sequential compressor as a representative example of shallow models, the node ordering should be cheap to establish from a random ordering, preferably in linear time.

Since we are testing the capacity of the general purpose compressor to perform inference despite the problem of isomorphism, we will actually present it with a worst case scenario. We will use a random ordering of nodes for all graphs. If the general purpose compressor still outperforms the random baseline under these conditions, it will tell us that it is, at least in part, resistant to the problem of isomorphism.

@@

The first is the classic Erd\H{o}s-R\'enyi (ER) model, where a uniform random choice is made from all graphs with $n$ nodes and $m$ links. The second is the Barab\'asi-Albert (BA) model \cite{albert2002statistical}, which grows a graph from a set of $n_0$ unconnected nodes, one node at a time, connecting each new node to $k$ distinct existing nodes where the probability that a given existing node is chosen for a connection is its degree, divided by the sum of the degrees of all nodes. Thus, under the BA model the more links a node has, the higher the probability that it will accrue even more. This effect causes the degree distribution of a BA network to become scale-free (ie. it follows a power law).

@@

To assess the performance of the clustering we label the clusters so that the accuracy is maximized (essentially assigning the optimal labels). Clearly, this would be cheating when testing a classifier, but since we are only interested in the clustering aspect, it gives us a straightforward performance measure.

@@

Subdue \cite{ketkar2005subdue} is an algorithm for finding frequent subgraphs in graph data. The algorithm searches for the subgraph that maximally compresses the data. The body of the algorithm is essentially a beam search through the space of subgraphs. It consists of three main routines:

\begin{description}
\item[Subgraph matching] This is an algorithm for finding the occurrences of a given subgraph in a graph. The method used is detailed in \cite{bunke1983inexact}. Since this is  a semi-exhaustive search for the solution to the NP-complete problem of subgraph isomorphism, the matching can only be solved for very small subgraphs. Unfortunately, even with subgraphs of four or five nodes, the matching is too slow in combination with the number of times it is executed to calculate a full distance matrix. To combat this issue, we remove all but the first $b_{\mbox{inner}}$ elements from the search queue after it is sorted at ech iteration, effectively turning the algorithm into a beam search.
\item[MDL Scoring]
This routine takes a subgraph, finds its occurrences in the data by the previous routine and replaces these occurrences with a single node with a new label, which we call a \emph{symbol node}. All links that connected to nodes in the subgraph are now connected to the symbol node. The links connecting to the symbol node are annotated with the indices of the nodes in the subgraph they were connected to. Storing the subgraph once together with the reduced data graph is enough to reconstruct the original data, and, if the subgraph occurs often, will lead to a significant compression.
\item[Subgraph search]
This is the `outer loop' of the algorithm. Starting with a graph of a single link between two nodes, it searches through the space of all connected graphs by extending each current candidate by one link at a time (possibly by adding a new node as well). The buffer of current candidates is sorted by MDL score, and the candidate with the highest score is extended to create new candidates. At each iteration all but the top $b_{\mbox{outer}}$ candidates are removed, turning the search into a beam search.
\end{description}

@@

	\hspace*{5mm}\textit{sort $S$ by} score($s', G$) \textit{for} $s' \in S$\\
	\hspace*{5mm}\textit{remove all but the first $b$ elements of $S$} \\
	\\
	\textbf{function} score$(s', G)$\\
	\hspace*{5mm} \textit{replace occurrences of $s'$ in $G$ with node $N$}\\
	\hspace*{5mm} \textit{annotate links to $N$ with the nodes in $s'$}\\
	\hspace*{5mm} \textbf{return} \textit{nr of bits to store the edited $G$ and $s'$ }\\
}
	\caption{Pseudocode for the Subdue algorithm}

@@

Table \ref{table:synthetic-large} shows the results on randomly generated graphs. Table \ref{table:real-life} shows the results for subgraphs sampled from real-life datasets.

@@

As for the graph-compressors, we see a small improvement relative to the sequential compressors for a strong increase in computational resources. Subdue as used in this paper is a relatively simple compressor, which isolates only a single subgraph for compression and we tested it only at modest parameters. The publications surrounding Subdue offer much more complex solutions (such as the induction of graph grammars \cite{jonyer2004mdl}). To investigate the promise of these models as compressors further, it will be necessary to investigate both parallelized versions of these algorithms and a more elegant relaxation of the exhaustive nature of their components.

@@

The method of coding data is always a sensitive point in compression-based learning. The precise choices made in translating the data to a string of bits can greatly affect which patterns are picked up, or ignored by the subsequent inference procedure. Here we detail the procedure used to encode the graphs in both compressors.

\subsection*{GZIP}
We take half of the adjacency matrix ($(n^2 + n)/2$ bits), and store it in flattened form as an array of java byte primitives, together with the size of the string (since the stored representation is padded to a multiple of eight). We then serialize this representation into a java GZIPOutputStream. To make our implementation generic, we assume that the nodes and links are labeled, and serialize the labels in a fixed order after the adjacency matrix. In the graphs mentioned in this paper, there is a single label assigned to all elements, so inference shouldn't be affected by the labels.

@@

To store a plain graph, we follow roughly the coding strategy outlined in \npcite{holder1994substructure}.

We first store the number of nodes $n$ in preefirx free coding, and the maximum number of neighbours for a node in the graph $n_{\mbox{max}}$ (int $\log n$ bits). We then store the lower half of the adjacency matrix (including the diagonal) row by row.

@@

We then account for the possibility of multiple links between node pairs. Since we expect miultiple links to be unlikely, we want to ensure that we do not waste many bits coding for this occasion if it does not occur. We store the pairs for which there are multiple links as a list consisting of two numbers per element: the first is the index of the pair in a canonical ordering, the second is the number of extra links that exist for that pair (in prefix-coding). We use a single bit to indicate whether this list will follow, and we reserve a codeword to signal the end of the list.

After this, all that remains is to encode the labels of the nodes and links. We assume that the sender and receiver in our coding scheme possess a codebook that is efficient for the data given, so that if a label $l$ occurs with frequency $\#l$, we can encode it in $-\log \frac{\#l}{\sum_k \#k}$ bits.

We assume that there is some canonical ordering among the nodes and links, and store them as a stream. Since the number of labels is known from the adjacency matrix, the code for the entire graph is self-delimiting.

@@

Finally, we must store for every link leading into a symbol node which node in the substructure it should connect to to restore the original data. Since we know the graph and all the symbol nodes, we can presume there to be a canonical ordering in these links, and we use $\log n_s$ bits for each, where $n_s$ is the size of the substructure in nodes.
